---
title: "01-data-compilation"
format:
  html:
    code-fold: show
    code-tools: true
    code-copy: true
    toc: true
    toc-location: left
    number-sections: true
    number-depth: 2
editor: source
---

# Set-up

```{r, warning=FALSE}
# | warning: false
# | message: false
library(tidyverse)
library(patchwork)
library(svglite)
library(scales)
library(treemapify)
library(compositions); library(vegan)
library(clustMixType)
library(lunar)
```

# Preliminary input data

All data from Port Aransas. Counts by hour. Remove data before January 1 2008, as it was newly deployed and has some inconsistencies. Also using lubridate to convert to R-friendly date/time formats. Note that Hurricane Harvey destroyed the pier housing the IFCB, so data stop right when Harvey was hitting the coastline.

Time zone is UTC and units of carbon are pg C per cell.

```{r}
# head(counts)
counts <- read.csv("input-ifcb-data/counts_hourly_Hu.csv") |>  
  # Use lubridate package to assign "date" format.
  mutate(DAY = ymd_hms(X)) %>% 
  mutate(DATE = date(DAY),
         MONTH = month(DAY),
         HOUR = hour(DAY),
         MIN = minute(DAY)) %>% 
  select(DAY, DATE, MONTH, HOUR, MIN, everything(), -X) %>% 
  filter(as_date(DAY) >= as_date("2008-01-01") & 
           as_date(DAY) <= as_date("2017-08-24"))
# glimpse(counts)
# range(counts$DAY)
# head(counts)
```

Carbon biomass by day. pgC per cell.

```{r}
carbon <- read.csv("input-ifcb-data/carbon_hourly.csv") %>% 
  mutate(DAY = ymd_hms(X)) %>% 
  mutate(DATE = date(DAY),
         MONTH = month(DAY),
         HOUR = hour(DAY),
         MIN = minute(DAY)) %>% 
  select(DAY, DATE, MONTH, HOUR, MIN, everything(), -X) %>% 
  filter(as_date(DAY) >= as_date("2008-01-01") & 
           as_date(DAY) <= as_date("2017-08-24"))
# head(carbon)

range(carbon$DAY)
# glimpse(carbon)
```


Subset so dataset includes "2008-01-01" to "2017-08-24".

```{r}
head(carbon)
head(counts)
```

### Export taxa names for manual curation

Export for manual curation

```{r}
# tax_export <- counts %>% 
#     pivot_longer(cols = -c(DAY), names_to = "CELL_ID", values_to = "COUNT") %>% #Add in time here when we have it.
#     select(CELL_ID) %>% distinct()
```

### Add taxa curated to cell IDs and carbon data

Siddarth contributed to this taxa list.

```{r}
tax_curated <- read.csv("input-ifcb-data/tax-list.csv") |>
  select(CELL_ID:Species) 
```

Add to cell ID information

```{r}
counts_long_wtax <- counts %>% 
  pivot_longer(cols = -c(DAY, DATE, MONTH, HOUR, MIN), names_to = "CELL_ID", values_to = "COUNT", values_drop_na = TRUE) %>% 
  filter(COUNT > 0) %>% 
  left_join(tax_curated)

head(counts_long_wtax)
# range(counts_long_wtax$COUNT)
```

Add to carbon data

```{r}
carbon_long_wtax <- carbon %>% 
  pivot_longer(cols = -c(DAY, DATE, MONTH, HOUR, MIN), names_to = "CELL_ID", values_to = "CARBON", values_drop_na = TRUE) %>% 
  filter(CARBON > 0) %>% 
  left_join(tax_curated)

# head(carbon_long_wtax)
# range(carbon_long_wtax$CARBON)
```

```{r}
save(carbon_long_wtax, counts_long_wtax, file = "output-data/counts-carbon-data-ifcb.RData")
```


## Combine carbon and cell counts

_Optional_
```{r}
combined_gom_long <- carbon_long_wtax %>% 
  left_join(counts_long_wtax)
```

Types of classifications in dataset.

```{r}
table(combined_gom_long$COURSE_CLASSIFICATION)
table(combined_gom_long$TYPE)
# glimpse(combined_gom_long)
# table(combined_gom_long$COURSE_CLASSIFICATION)
```

> What is the range of cellular carbon for each cell ID?

There are 130 entries.

```{r}
length(unique(combined_gom_long$CELL_ID))
# unique(combined_gom_long$Phylum)
```

```{r, fig.height=7, fig.width=28}
rm <- c("IFCB excess", "Larvae")
# hist(combined_gom_long$CARBON)
combined_gom_long %>% 
  # filter(!(COURSE_CLASSIFICATION %in% rm)) %>% 
  filter(COURSE_CLASSIFICATION == "Microeukaryote") %>%
  # filter(Supergroup == "Alveolata") %>% 
  ggplot(aes(x = CELL_ID, y = CARBON, fill = Phylum)) +
  geom_violin() +
  facet_grid(cols = vars(Phylum), scales = "free_x", space = "free_x") +
  # facet_wrap(vars(Phylum), scales = "free", as.table = FALSE) +
  theme_classic() +
  scale_y_log10() +
  labs(x = "", y = bquote("pg C"~cell^-1)) +
  theme(legend.position = "bottom", legend.title = element_blank(),
        strip.text.y = element_text(color = "black", face = "bold", size = 14),
        strip.background = element_blank(),
        strip.placement = "outside",
        axis.text.x = element_text(color = "black", size = 10, angle = 45, vjust = 1, hjust = 1),
        axis.title = element_text(color = "black", size = 10),
        axis.text.y = element_text(color = "black", size = 10),
        panel.grid.major = element_line(color = "#d9d9d9", linewidth = 0.2))
# ?facet_wrap()
```


# Seasonal definition

Make data frames to assign season

Spring = March 1 - May 31 Summer = June 1 - Aug 31 Fall = Sept 1 - Nov 30 Winter = Dec 1 - Feb (March 1 (-1))

Make a new dataframe to just add seasons (manually) to the date. I'm subsetting to 2024 to include leap day years.

```{r}
# df_season <- counts %>% 
#   pivot_longer(cols = -c(DAY, DATE, MONTH, HOUR), names_to = "CELL_ID", values_to = "COUNT") %>%
#   mutate(DATA_GAP = case_when(
#     is.na(COUNT) ~ "Missing",
#     TRUE ~ "Not missing"
#   )) |> 
#   mutate(date_tmp =  as.Date(format(DAY, '2024-%m-%d')), 
#         SEASON = case_when(between(date_tmp, ymd('2024-03-01'), ymd('2024-05-31')) ~ "Spring",
#                            between(date_tmp, ymd('2024-06-01'), ymd('2024-08-31')) ~ "Summer", 
#                            between(date_tmp, ymd('2024-09-01'), ymd('2024-11-30')) ~ "Fall", 
#                            TRUE ~ "Winter")) %>% # Winter needs to be "else" because you can't select between.
#   select(DAY, date_tmp, SEASON, DATA_GAP) |> distinct() |> 
#   mutate(YEAR = year(DAY),
#          MONTH = month(DAY),
#          DATE = day(DAY),
#          HOUR = hour(DAY)) |> 
#   group_by(SEASON) |> 
#   mutate(month_start_tmp = case_when(
#     SEASON == "Winter" ~ max(MONTH),
#     TRUE ~ min(MONTH)),
#     year_start_tmp = case_when(SEASON == "Winter" ~ YEAR - 1,
#                                TRUE ~ YEAR),
#     ) |> 
#   mutate(month_end_tmp = case_when(
#     SEASON == "Winter" ~ 2,
#     TRUE ~ max(MONTH))) |> 
#   group_by(SEASON, MONTH) |> 
#   mutate(day_start_tmp = min(day(DAY)),
#          day_end_tmp = case_when(
#            (SEASON == "Winter" & leap_year(YEAR) == TRUE) ~ 29,
#            (SEASON == "Winter" & leap_year(YEAR) == FALSE) ~ 28,
#            (SEASON == "Summer" | SEASON == "Spring") ~ 31,
#            SEASON == "Fall" ~ 30
#          )) |> 
#   ungroup() |> 
#   unite("start", year_start_tmp, month_start_tmp, day_start_tmp, sep = "-", remove = FALSE) |> 
#   unite("end", YEAR, month_end_tmp, day_end_tmp, sep = "-", remove = FALSE) |> 
#   mutate(START = ymd(start),
#          END = ymd(end),
#          START_noy = mdy(paste(month(START), day(START), "2024", sep = "-")),
#          END_noy = mdy(paste(month(END), day(END), "2024", sep = "-"))) |> 
#   select(DAY, DATE, MONTH, YEAR, HOUR, SEASON, START, END, START_noy, END_noy) |> distinct()
# 
# # head(df_season)
# # View(df_season |> select(SEASON, START, END) |> distinct())
# df_oneyear <- df_season |> 
#   mutate(DATE = mdy(paste(month(DAY), day(DAY), "2024", sep = "-"))) |> 
#   select(SEASON, START_noy, END_noy, DATE) |> 
#   # filter(!(SEASON == "Winter" & (month(START_noy) == 12))) |> 
#   distinct()
# # head(df_oneyear)
# 

```

Re-evaluate seasons
```{r}
seasons <- tribble(~season, ~start_doy, ~end_doy,
  "Cool", 340, 113,
  "Spring freshening",114, 182,
  "Hot hypersaline", 183, 251,
  "Autumn low-inflow", 252, 339) |>
  mutate(
    start_date = format(as_date("2001-01-01") + (start_doy - 1), "%b %d"),
    end_date = format(as_date("2001-01-01") + (end_doy - 1),   "%b %d"),
    dates = str_c(start_date, " - ", end_date),
    n_days = if_else(end_doy >= start_doy,
                     end_doy - start_doy + 1,
                     365 - start_doy + 1 + end_doy))

```

function to assign season
```{r}
assign_season <- function(doy) {
  case_when(
    doy >= 340 | doy <= 113 ~ "Cool",
    doy <= 182 ~ "Spring freshening",
    doy <= 251~ "Hot hypersaline",
    TRUE ~ "Autumn low-inflow")
}
```

```{r}
head(counts_long_wtax)
```


Add seasons to count and carbon data
```{r}
counts_seasonal <- counts_long_wtax %>% 
  mutate(DATA_GAP = case_when(
    is.na(COUNT) ~ "Missing",
    TRUE ~ "Not missing"
  )) |> 
  mutate(YEAR = year(DAY),
         doy = yday(DAY)) |> 
  mutate(SEASON = assign_season(doy))
# unique(counts_seasonal$doy)
unique(counts_seasonal$SEASON)


carbon_seasonal <- carbon_long_wtax %>% 
  mutate(DATA_GAP = case_when(
    is.na(CARBON) ~ "Missing",
    TRUE ~ "Not missing"
  )) |> 
  mutate(YEAR = year(DAY),
         doy = yday(DAY)) |> 
  mutate(SEASON = assign_season(doy))
```

**Seasons**

* _Cool_ : Dec-Feb, cold fronts add low temperatures and stronger mixing from wind. 
* _Spring freshening_ : Mar-June, start seeing increased rain, and therefore lower salinity values, increased nutrients and increased riverine height.
* _Hot Hypersaline_ : Lower inflow from rivers, drought-risk season, high evaporation, warm, stratified conditions start emerging. Increased salinity. July-sept
* _Autumn low-inflow_ : Oct-Nov, increased salinity continues into the Fall, but temperatures start declining. water column usually remains stratefied, some major storms will arise and add punctuated mixing.

```{r}
# counts_seasonal
```

# Add metadata

See script `environmental-data-compilation.qmd` to gather all environmental data.

```{r}
load("input-env-data/metadata-09202026.RData", verbose = TRUE)
```

> Confirm QC here for all metadata

```{r}
str(metadata_storms_env)
# unique(metadata_storms_env$MIN)
```
```{r}
head(counts_seasonal)
unique(counts_seasonal$HOUR)
```

```{r}
head(metadata_storms_env)
unique(metadata_storms_env$HOUR)
```


```{r}
dim(metadata_storms_env); dim(counts_seasonal)

re_df <- metadata_storms_env |>
  select(-any_of(c("DAY", "MONTH", "MIN"))) |>
  mutate(DATE = as_date(DATE), HOUR = as.integer(HOUR))

re_df |> count(DATE, HOUR) |> filter(n > 1)     # should return nothing

counts_seasonal_metadata <- counts_seasonal |>
  mutate(HOUR = as.integer(HOUR)) |>
  left_join(re_df, by = c("DATE", "HOUR"))
# dim(counts_seasonal); dim(counts_seasonal_metadata) # Should be equal

carbon_seasonal_metadata <- carbon_seasonal |>
  mutate(HOUR = as.integer(HOUR)) |>
  left_join(re_df, by = c("DATE", "HOUR"))
# dim(carbon_seasonal); dim(carbon_seasonal_metadata)

# stopifnot(nrow(counts_seasonal_metadata) == nrow(counts_seasonal))
```

```{r}
counts_seasonal_metadata |> count(HOUR) |> print(n = 24)   # expect all 24 hours
sum(is.na(counts_seasonal_metadata$tide_direction))         # expect a small number
```


## QC tide height & direction

First Exploring tide height

```{r}
counts_seasonal_metadata %>%
  group_by(MLLW, tide_direction, DIFFERENCE) %>%
  summarize(COUNT_SUM = sum(COUNT)) %>%
  ggplot(aes(x = DIFFERENCE, y = COUNT_SUM, fill = tide_direction)) +
  geom_point(shape = 21) +
  scale_y_log10()
```


This is how we are "de-tiding".

```{r}
counts_seasonal_metadata_clean <- counts_seasonal_metadata %>%
  filter(tide_direction != "Outgoing")

carbon_seasonal_metadata_clean <- carbon_seasonal_metadata %>%
  filter(tide_direction != "Outgoing")
```

### Save all data
```{r}
save(seasons, counts_seasonal_metadata_clean, carbon_seasonal_metadata_clean, file = "ifcb-data-compiled_09202026.RData")
```


# Map

```{r}
# install.packages(c("googleway", "ggrepel", 
# "ggspatial", "libwgeom", "sf", "rnaturalearth", "rnaturalearthdata"))

library(rnaturalearth)
library(rnaturalearthdata)
library(sf)

world <- ne_countries(scale='medium',returnclass = 'sf')
class(world)
```

## Map of whole Gulf region

```{r}
ggulf <- ggplot(data = world) +
  geom_sf(aes(fill = region_wb)) +
  annotate(geom = "text", x = -90, y = 26, label = "", 
     fontface = "italic", color = "grey22", size = 6) +
  coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE) +
  scale_fill_viridis_d(option = "plasma") +
  theme(legend.position = "none", axis.title.x = element_blank(), 
     axis.title.y = element_blank(), panel.background = element_rect(fill = "azure"), 
     panel.border = element_rect(fill = NA))
ggulf
```

```{r, fig.width=7, fig.height=7}
# MAP
ggplot(data = world) +
  geom_sf(color = "black", linewidth = 0.6, fill = "grey10", alpha = 0.8) +
  # Region of study
  coord_sf(xlim = c(-100, -80), ylim = c(22, 32), expand = FALSE) +
  #
  # Add data points for stations
  # annotate("point", x = -94.8, y = 29.3, size = 10, shape = 21, color = "red", fill = "red") +
  # geom_point(aes(x = -94.8, y = 29.3), shape = 21, size = 5) +
  # geom_rect(xmin = -93.4, xmax = -93.7, ymin = 27.5, ymax = 27.9, fill = "pink", color = "grey10", alpha = 0.5) + #inset
  labs(y = "Latitude (°N)", x = "Longitude (°W)", fill = "") +
  scale_x_reverse() +
  theme_classic() +
  theme(legend.position = "none",
    axis.text = element_text(size = 10, colour="black"),
        panel.background = element_rect(fill = "#B4C7D9"),
        panel.border = element_rect(fill = NA))
```


# Session Info
```{r}
sessionInfo()
```

