---
title: "02-overview-seasonality-stats"
format:
  html:
    code-fold: show
    code-tools: true
    code-copy: true
    toc: true
    toc-location: left
    number-sections: true
    number-depth: 2
editor: source
---

# Set up


Run after `01-data-compilation.qmd`

```{r, warning=FALSE}
# | warning: false
# | message: false
library(tidyverse)
library(patchwork)
library(svglite)
library(scales)
library(treemapify)
library(compositions); library(vegan)
library(clustMixType)
library(lunar)
```

```{r}
load("ifcb-data-compiled_09202026.RData", verbose = TRUE)
```

```{r}
head(counts_seasonal_metadata_clean)
```


# Seasonal classification & anomaly check

Get stats on parameters for each season. 

```{r}
min_hours_day <- 1 # hourly records required to keep a daily mean
window_days <- 15 # +/- days around each day of year for the climatology
min_ref_days <- 30 # minimum reference days needed to score a day
z_cut <- 3.5  # (abs(robust z)) above this = anomalous
pct_lo <- 0.05 # seasonal lower threshold percentile
pct_hi <- 0.95   # seasonal upper threshold percentile
anomaly_vars <- c("SAL", "gage_height_ft", "TEMP", "discharge_cfs") # any of TEMP, SAL, gage_height_ft, discharge_cfs
```

```{r}
head(seasons)
```


```{r}
thresholds <- counts_seasonal_metadata_clean |>
  select(season = SEASON, TEMP, SAL, PRECIP_in, gage_height_ft, discharge_cfs) |> 
  distinct() |> 
  pivot_longer(cols = -c(season)) |> 
  drop_na(value) |> 
  group_by(season, name) |>
  summarise(
    n_days = n(), #Number of days
    min = min(value), # min value
    mean = mean(value),
    max = max(value),
    p05 = quantile(value, pct_lo), # percent low threshold 
    p95 = quantile(value, pct_hi), # percent high threshold
    .groups = "drop") |>
  left_join(select(seasons, season, 
                   dates, start_doy, 
                   end_doy), by = "season") |>
  select(season, dates, start_doy, 
         end_doy, name, n_days, 
         min, p05, mean, p95, max) |>
  arrange(name, season)
```

```{r}
thresholds
```

Quick table environmental parameters.
```{r}
thresholds_wide <- thresholds |>
  mutate(across(c(min, p05, mean, p95, max), ~ signif(.x, 4))) |>
  mutate(range = str_c(min, " / ", mean, " / ", max,
                       "  (", p05, "-", p95, ")")) |>
  select(season, dates, name, range) |>
  pivot_wider(names_from = name, values_from = range)

thresholds_wide
```

### Storm summary

```{r}
head(counts_seasonal_metadata_clean)
# unique(counts_seasonal_metadata_clean$storm_id)
unique(counts_seasonal_metadata_clean$storm_phase)
```


```{r}
storm_sum <- counts_seasonal_metadata_clean |> 
  filter(storm_phase == "storm") |> 
  # select(storm_end_utc, storm_start_utc)
  mutate(storm_duration = ceiling(as.numeric((ymd_hms(storm_end_utc) - ymd_hms(storm_start_utc)), "days"))) |> 
  select(season = SEASON, storm_id, storm_type, tropical_system, severity_score, storm_duration) |> 
  distinct() |> 
  group_by(season, storm_type) |> 
  summarize(NUM_OF_STORMS = n(),
            SEVERITY_MIN = min(severity_score),
            SEVERITY_MAX = max(severity_score),
            SEVERITY_MEAN = mean(severity_score),
            STORM_LENGTH_MIN = min(storm_duration),
            STORM_LENGTH_MAX = max(storm_duration),
            STORM_LENGTH_MEAN = mean(storm_duration))
storm_sum

# unique(storm_sum$tropical_system)
```

# Overview figure: annual seasonality of environmental parameters

One-year composite (day of year on the x-axis) of the environmental record. Each line is the multi-year mean; the ribbon is the spread across years. Points are storm events (start date), placed at that year's value. Dashed lines connect across weeks with too few years of data.

## Parameters

```{r}
bin_days <- 7 # width of each day-of-year bin (days)
min_hours_bin <- 24 # minimum sampled hours in a year x bin to keep it (IFCB sampling is not continuous)
min_years_bin <- 3 # minimum distinct years required to report a bin mean
band_type <- "sd" # spread across years shown as ribbon: "sd" (std deviation) or "se" (std error of the mean)
env_labels <- c(TEMP = "Temperature (\u00b0C)",
                SAL = "Salinity",
                PRECIP_in = "Precipitation (in per day)",
                gage_height_ft = "Gauge height (ft)",
                discharge_cfs = "Discharge (1000 cfs)",
                MLLW = "Water level (m MLLW)") # panel order = order here
```

```{r}
month_starts <- yday(as_date(str_c("2001-", 1:12, "-01"))) # doy of the first of each month (non-leap year)
storm_pal <- c("Severe convective" = "#c19f2c",
               "Flood / heavy rain" = "#d96629",
               "Cold front / norther" = "#6450a1",
               "Coastal flood / swell" = "#009E73",
               "Tropical cyclone" = "#642d5e") # colourblind-safe, one per storm type
```

## Hourly to weekly bins by year

Hourly environmental records are collapsed to one value per year and day-of-year bin. All parameters are averaged over the sampled hours; precipitation is the mean hourly rate scaled to inches per day (IFCB sampling is not continuous, so a sum over sampled hours would be biased).

```{r}
env_hourly <- counts_seasonal_metadata_clean |>
  select(DAY, YEAR, doy, all_of(names(env_labels))) |>
  distinct() |> # counts are in long format, so collapse to one row per hour
  mutate(doy = pmin(doy, 365), # fold leap day into Dec 31
         bin = (doy - 1) %/% bin_days + 1, # 1 = Jan 1-7, etc.
         discharge_cfs = discharge_cfs / 1000)
head(env_hourly)
```

```{r}
env_year_bin <- env_hourly |>
  pivot_longer(cols = all_of(names(env_labels)), names_to = "name", values_to = "value") |>
  drop_na(value) |>
  group_by(name, YEAR, bin) |>
  summarise(value = mean(value),
            n_hours = n(),
            .groups = "drop") |>
  mutate(value = if_else(name == "PRECIP_in", value * 24, value)) |> # in per hour -> in per day
  filter(n_hours >= min_hours_bin) # drop thinly sampled year x bin combinations
head(env_year_bin)
```

## Mean and spread across years

```{r}
env_clim <- env_year_bin |>
  group_by(name, bin) |>
  summarise(n_years = n_distinct(YEAR),
            mean = mean(value),
            sd = sd(value),
            .groups = "drop") |>
  mutate(se = sd / sqrt(n_years),
         spread = if (band_type == "se") se else sd) |>
  right_join(expand_grid(name = names(env_labels), bin = 1:ceiling(365 / bin_days)),
             by = c("name", "bin")) |> # add empty rows so gaps are explicit
  mutate(missing = is.na(mean) | n_years < min_years_bin,
         across(c(mean, sd, se, spread), ~ if_else(missing, NA_real_, .x)),
         x_doy = (bin - 1) * bin_days + (bin_days + 1) / 2, # bin midpoint
         name = factor(name, levels = names(env_labels)))
env_clim |> filter(missing) |> count(name)
```

```{r}
write.csv(env_clim, file = "output-data/env-annual-climatology.csv", row.names = FALSE)
```

## Storm events

Each storm is placed at its start date, at that year's binned value of each parameter.

```{r}
storm_events <- counts_seasonal_metadata_clean |>
  filter(storm_phase == "storm") |>
  transmute(storm_id, storm_type, YEAR, doy = pmin(doy, 365)) |>
  group_by(storm_id, storm_type, YEAR) |>
  summarise(doy = min(doy), .groups = "drop") |> # first day of the storm
  mutate(bin = (doy - 1) %/% bin_days + 1,
         x_doy = doy)
head(storm_events)
```

```{r}
storm_points <- storm_events |>
  inner_join(env_year_bin, by = c("YEAR", "bin"), relationship = "many-to-many") |> # one point per storm per parameter
  mutate(name = factor(name, levels = names(env_labels)),
         storm_type = factor(storm_type, levels = names(storm_pal)))
storm_points |> distinct(storm_id, storm_type) |> count(storm_type)
```

## Season shading

Season boundaries come from the `seasons` table. Cool wraps the new year, so it is split into two rectangles.

```{r}
season_rects <- seasons |>
  mutate(xmin = list(start_doy), xmax = list(end_doy)) |>
  select(season, start_doy, end_doy) |>
  pmap_dfr(function(season, start_doy, end_doy) {
    if (end_doy >= start_doy) {
      tibble(season, xmin = start_doy - 0.5, xmax = end_doy + 0.5)
    } else {
      tibble(season, xmin = c(0.5, start_doy - 0.5), xmax = c(end_doy + 0.5, 365.5)) # wraps the year
    }
  }) |>
  mutate(shade = season %in% c("Spring freshening", "Autumn low-inflow")) # alternate shading
season_rects
```

## Plot environmental data

```{r}
season_pal <- c("Cool" = "#5d4e62", 
                "Spring freshening" = "#bccbae",
                "Hot hypersaline" = "#cc5500", 
                "Autumn low-inflow" = "#ffbf00")
```
 
 
### Season color bar

A single bar spanning the same day-of-year axis as the panels below.

```{r}
season_bar <- ggplot(season_rects) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1, fill = season),
            colour = "white", linewidth = 0.4) +
  # geom_text(data = season_rects |> group_by(season) |> slice_max(xmax - xmin, n = 1), # label the longest piece of a wrapped season
            # aes(x = (xmin + xmax) / 2, y = 0.5, label = season),
            # colour = "white", fontface = "bold", size = 3.6) +
  scale_x_continuous(limits = c(0.5, 365.5), expand = c(0, 0)) +
  scale_fill_manual(values = season_pal, guide = "none") +
  theme_void()
```

Solid lines are runs of consecutive bins; the dashed line bridges bins with missing data.

```{r}
env_runs <- env_clim |>
  arrange(name, bin) |>
  group_by(name) |>
  mutate(run = cumsum(missing)) |> # a new run starts after each missing bin
  ungroup() |>
  filter(!missing)
```

### Define environmental panels

```{r}
env_panels <- ggplot() +
  geom_rect(data = filter(season_rects, shade),
            aes(xmin = xmin, xmax = xmax, 
                ymin = -Inf, ymax = Inf),
            fill = "grey92") +
  # RIBBON for spread
  geom_ribbon(data = env_runs,
              aes(x = x_doy, ymin = if_else(name == "PRECIP_in", pmax(mean - spread, 0), mean - spread), # stop at 0
                  ymax = mean + spread,
                  group = interaction(name, run)),
              fill = "grey20", alpha = 0.4) +
  # Get NAs filled in with dash. shows gaps
  geom_line(data = env_runs, 
            aes(x = x_doy, y = mean),
            linetype = "dashed", 
            color = "grey30", linewidth = 0.4) +
  geom_line(data = env_runs, aes(x = x_doy, y = mean, group = interaction(name, run)),
            color = "black", linewidth = 0.7) + # solid where consecutive bins exist
  geom_point(data = storm_points,
             aes(x = x_doy, y = value, fill = storm_type),
             shape = 21, color = "white", stroke = 0.4, size = 3, alpha = 0.9) +
  facet_grid(name ~ ., scales = "free_y", switch = "y",
             labeller = labeller(name = env_labels)) +
  scale_x_continuous(breaks = month_starts + 14, labels = month.abb,
                     limits = c(0.5, 365.5), expand = c(0, 0)) +
  scale_fill_manual(values = storm_pal, name = "Storm type", drop = FALSE) +
  guides(fill = guide_legend(nrow = 2, override.aes = list(size = 3))) +
  labs(x = "", y = "") +
  theme_bw(base_size = 12) +
  theme(strip.placement = "outside",
        strip.background = element_blank(),
        strip.text.y.left = element_text(angle = 90),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.spacing.y = unit(0.4, "lines"),
        legend.position = "bottom")
```


### Stack

Stack the season bar on top of the panels.

```{r}
#| fig-height: 11
#| fig-width: 9
env_plot <- season_bar / env_panels + plot_layout(heights = c(0.05, 1))
env_plot
```
Parameters: storm type is shown. The line is the mean across all 9 years of data, binned (averaged) by 7 day increments. And ribbon around the line is showing the standard deviation across years. Dashed lines are connecting where there is less than 3 years of data ,usually a big missing gap.


# Seasonal anomalies

```{r}
circ_dist <- function(a, b, n = 366) pmin(abs(a - b), n - abs(a - b))
```

Pool each day of year across all years and sets a window number of days to consider. 

```{r}
# climatology <- function(dat) {
#   # dat: doy, value for one variable
#   map_dfr(1:366, function(d) {
#     ref <- dat$value[circ_dist(dat$doy, d) <= window_days]
#     tibble(doy = d,
#            n_ref = length(ref),
#            clim_median = median(ref),
#            clim_mad = 1.4826 * mad(ref, constant = 1))
#   })
# }

min_ref_years <- 4 # minimum distinct other years in the window to score a day

climatology <- function(dat) {
  # dat: year, doy, value for one variable; each year is scored against the other years
  expand_grid(year = unique(dat$year), doy = 1:366) |>
    mutate(stats = map2(year, doy, \(y, d) {
      in_ref <- dat$year != y & circ_dist(dat$doy, d) <= window_days
      ref <- dat$value[in_ref]
      tibble(n_ref = length(ref),
             n_ref_years = n_distinct(dat$year[in_ref]),
             clim_median = median(ref),
             clim_mad = 1.4826 * mad(ref, constant = 1))
    })) |>
    unnest(stats)
}
```

Score each parameters based on a z score within each season.
```{r}
scored <- counts_seasonal_metadata_clean |>
  select(season = SEASON, DATE, doy, year = YEAR, TEMP, SAL, PRECIP_in, gage_height_ft, discharge_cfs) |> 
  distinct() |> 
  pivot_longer(cols = -c(season, DATE, year, doy)) |> 
  drop_na(value) |> 
  filter(name %in% anomaly_vars) |>
  group_by(name) |>
  group_modify(function(dat, key) {
    clim <- climatology(dat)
      dat |>
        left_join(clim, by = c("year", "doy")) |>
  mutate(robust_z = if_else(n_ref_years >= min_ref_years & clim_mad > 0,
                            (value - clim_median) / clim_mad, NA_real_))
      }) |>
  ungroup() |>
  left_join(select(thresholds, season, name, p05, p95),
            by = c("season", "name")) |>
  mutate(flag_z = abs(robust_z) > z_cut,
    flag_seasonal = value < p05 | value > p95,
    direction = case_when(
      robust_z > 0 ~ "high", robust_z < 0 ~ "low",
      TRUE ~ NA_character_))

```

* Climate median and mad represent the typical spread for the values in question. The MAD is the median absolute deviation.
* robust z is a z-score, with (value - clim_median)/climate_mad, it considers the number of days of a specific value. 
* when the value is outside the p05-p95 value, it is noted as "TRUE" in flag seasonal and the direction shows if the value is higher or lower than the expected seasonal value.

To look for seasonal anomalies, subset for `flag_seasonal == TRUE`. 

```{r}
scored
```

Auto detection of anaomalous weather events
```{r}
events <- scored |>
  filter(flag_z, !is.na(direction)) |>
  arrange(name, direction, DATE) |>
  group_by(name, direction) |>
  mutate(event_id = cumsum(c(TRUE, diff(DATE) != 1))) |>
  group_by(name, direction, event_id) |>
  summarise(
    start = min(DATE),
    end  = max(DATE),
    days = n(),
    peak_value = value[which.max(abs(robust_z))],
    peak_z = round(robust_z[which.max(abs(robust_z))], 1),
    season = season[which.max(abs(robust_z))],
    days_outside_seasonal_range = sum(flag_seasonal),
    .groups = "drop"
  ) |>
  select(-event_id) |>
  arrange(name, start)
events
```

```{r}
season_pal <- c("Cool" = "#5d4e62", "Spring freshening" = "#bccbae",
                "Hot hypersaline" = "#cc5500", "Autumn low-inflow" = "#ffbf00")

anom_fig <- scored |>
  ggplot(aes(DATE, value)) +
  geom_line(colour = "grey70", linewidth = 0.3) +
  geom_point(data = filter(scored, flag_z),
             aes(colour = season), size = 1.6) +
  scale_colour_manual(values = season_pal, name = "Season of anomaly") +
  facet_wrap(~ name, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = NULL,
       title = "Anomalous days against a day-of-year climatology",
       subtitle = str_glue("Flagged where |robust z| > {z_cut}; daily means")) +
  theme_bw()

anom_fig
```

## Weather anomaly data

```{r}
save(scored, events, file = "output-data/weather-events.RData")
```


# Set seasonal blocks

## Data set up & parameters


```{r}
#| label: params
min_hours_detected <- 100 # a CELL_ID must be detected in at least this many hours to be analysed (set 0 to keep all)
min_obs_block <- 200      # minimum hourly observations for a season block to be analysed
min_span_h <- 240          # minimum span (hours) a block must cover; the periodogram scan is capped at span / 3
detrend_span <- 0.75      # loess span for removing the slow within-season trend
```


Builds a timestamp and keeps only the columns needed. `parse_number()` handles `HOUR` stored as text (e.g. "13:00"), and non-numeric `COUNT` values become `NA`.
 
```{r}
#| label: clean
hourly_long <- counts_seasonal_metadata_clean |>
  mutate(
    DATE = as_date(DATE),
    HOUR = parse_number(as.character(HOUR)),          # 0-23 as a number
    COUNT = suppressWarnings(as.numeric(COUNT)),
    datetime = as.POSIXct(DATE, tz = "UTC") + HOUR * 3600   # hourly timestamp
  ) |>
  filter(!is.na(datetime), !is.na(SEASON)) |>          # drop rows that failed to parse
  select(datetime, SEASON, CELL_ID, COUNT)
 
head(hourly_long)
```


Check for rows lost or unparsed dates before continuing.
 
```{r}
#| label: clean-check
c(rows_in = nrow(counts_seasonal_metadata_clean), rows_kept = nrow(hourly_long))
range(hourly_long$datetime)
```

# Define season blocks
 
Each uninterrupted run of one season in one year is a block. Blocks are analysed separately so that gaps and season boundaries never enter a spectrum, and so that repeats can be compared across years.

This is necessary for downstream periodicity, recurring trends analysis - because we need to detrend.

```{r}
#| label: sampled-hours
sampled_hours <- hourly_long |>
  distinct(datetime, SEASON) |>
  arrange(datetime)
```
 
```{r}
#| label: blocks
season_blocks <- sampled_hours |>
  mutate(BLOCK_ID = cumsum(SEASON != lag(SEASON, default = first(SEASON)))) |>  # new id at each season change
  group_by(BLOCK_ID) |>
  mutate(BLOCK_YEAR = year(min(datetime) + (max(datetime) - min(datetime)) / 2)) |>   # year of block midpoint
  ungroup()

season_blocks 
# season_blocks |> distinct(BLOCK_ID, SEASON, BLOCK_YEAR) |> count(SEASON, BLOCK_YEAR) |> filter(n > 1) #returns nothing
```


Using the *n_hours* parameter to determine minimum observations for season block to consider periodic time.

## Build the series
 
Every CELL_ID that passes the detection filter, plus a total-cells series for reference. Sampled hours with no detection are set to zero.
 
```{r}
#| label: top-taxa
top_taxa <- hourly_long |>
  group_by(CELL_ID) |>
  summarise(hours_detected = sum(COUNT > 0, na.rm = TRUE)) |>   # hours with at least one detection
  filter(hours_detected >= min_hours_detected) |>
  pull(CELL_ID)
 
length(top_taxa)
```

```{r}
#| label: series
total_series <- hourly_long |>
  group_by(datetime) |>
  summarise(COUNT = sum(COUNT, na.rm = TRUE)) |>
  mutate(CELL_ID = "All cells")
 
taxa_series <- sampled_hours |>
  select(datetime) |>
  cross_join(tibble(CELL_ID = top_taxa)) |>          # every sampled hour x every taxon
  left_join(
    hourly_long |> group_by(datetime, CELL_ID) |> 
      summarise(COUNT = sum(COUNT, na.rm = TRUE), .groups = "drop"),
    by = c("datetime", "CELL_ID")
  ) |>
  replace_na(list(COUNT = 0))                         # absent = zero
```


```{r}
#| label: series-combine
head(taxa_series) 
head(total_series)
count_series <- bind_rows(total_series, taxa_series)
# count_series
```


# Log-transform and detrend within each block
 
Log10(x + 1) tames the orders-of-magnitude range. The loess residual removes slow trends that would otherwise dominate the long-period end of the spectrum.
 
Defines a safe detrending function. It falls back to mean-centering when a series is too short or constant for loess.
 
```{r}
#| label: detrend-function
detrend_loess <- function(hrs, y, span) {
  ok <- is.finite(hrs) & is.finite(y)
  # If fewer than 20 observations, or fewer than 3 unique entires it will skip the detrended step.
  if (sum(ok) < 20 || length(unique(y[ok])) < 3) return(y - mean(y, na.rm = TRUE))
  fit <- loess(y[ok] ~ hrs[ok], span = span, degree = 1)
  out <- rep(NA_real_, length(y))
  out[ok] <- residuals(fit)
  out
}
```
 
Detrending step. This takes individual taxa and all the count data and performs a detrending step. Output is detrended log counts and there will be one block per taxa. Use this for periodicity and recurrence trend mining. 

```{r}
#| label: detrend
detrended <- count_series |>
  filter(!is.na(datetime)) |>
  inner_join(season_blocks, by = "datetime") |>
  group_by(CELL_ID, BLOCK_ID) |>
  filter(n() >= min_obs_block, # note from params
         as.numeric(difftime(max(datetime), min(datetime), units = "hours")) >= min_span_h) |>
  mutate(
    hrs = as.numeric(difftime(datetime, min(datetime, na.rm = TRUE), units = "hours")),
    # hrs = as.numeric(difftime(datetime, min(datetime), units = "hours")),  # hours since block start
    LOG_COUNT = log10(COUNT + 1),
    # fit a linear loess of the log count data
    ## See function above
    RESID = detrend_loess(hrs, LOG_COUNT, detrend_span)
  ) |>
  ungroup() |>
  filter(!is.na(RESID))
```

```{r}
head(detrended)
```

```{r}
#| label: block-summary
detrended |> distinct(CELL_ID, SEASON, BLOCK_YEAR, BLOCK_ID) |> count(SEASON, name = "n_taxa_blocks")
```

### Save detrended data
 
```{r}
#| label: save
write.csv(detrended, file = "output-data/hourly-season-blocks.csv", row.names = FALSE)
```

```{r}
# Unique years within each season (should return nothing)
detrended |> distinct(SEASON, BLOCK_ID, BLOCK_YEAR) |> count(SEASON, BLOCK_YEAR) |> filter(n > 1)

# Residuals centered and finite (mean near 0, no NA)
detrended |> group_by(CELL_ID, BLOCK_ID) |>
  summarise(m = mean(RESID), s = sd(RESID), .groups = "drop") |>
  summarise(max_abs_mean = max(abs(m)), n_flat = sum(s == 0), n_na = sum(is.na(s)))

# How sparse are the series? (prevalence = share of hours with a non-zero count)
detrended |> group_by(CELL_ID, BLOCK_ID) |>
  summarise(prev = mean(COUNT > 0), .groups = "drop") |>
  summarise(median_prev = median(prev), share_under_10pct = mean(prev < 0.10))
```


# Session info
```{r}
sessionInfo()
```

